---
title: "Example-4"
author: "Nathaniel J. Rhodes, PharmD, MSc and Paul R. Yarnold, PhD"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example-4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(root.dir= "data/Example 4")
```

## Example 4. Binary Class Variable, Ordered Attribute

Appleton studied the number of migraine attacks experienced in a clinical trial of two alternative treatments, for a sample of 67 patients.^[Appleton DR (1995). Pitfalls in the interpretation of studies: III. _Journal of the Royal Society of Medicine_, 88, 241-243.] Various conventional statistical methods were used to compare number of attacks between treatments, but all were compromised due to violations of underlying assumptions. _Student’s_ t-test was employed to compare the mean number of attacks between treatments, but wasn’t statistically significant _p_<0.14. The t-test was used again to compare the data after modification by a square root transformation _p_<0.06, and again after a log transformation _p_<0.07, but each of these analyeses failed to identify a reliable mean difference. The Mann-Whitney U-test was tried next _p_<0.07, but the requirements of the test weren’t met (i.e., there were too many tied values). A normal test assuming a Poisson distribution was then tried, and while it identified a statistically significant effect (p<0.04), the Poisson assumption was untenable.^1^ Finally, it was decided to spline the number of migraine attacks into a binary indicator variable:

>“…discretizing the data at some point, probably between no attacks and one or more” (p. 242)

Fisher’s exact test was used to validate (using a one-tailed test) the spline between 0 and 1 migraine attacks, exact _p_<0.0219.

In contradistiction to classical statistical methods, ODA is invariant over any monotonic transformation of the data^[Yarnold, P.R., & Soltysik, R.C. (2010). Optimal data analysis: A general statistical analysis paradigm. _Optimal Data Analysis_, 1, 10-22.], ODA is fundamentally non-parametric because the objective function is chance-corrected predictive accuracy,^[Yarnold, P.R., & Soltysik, R.C. (2016). _Maximizing Predictive Accuracy._ Chicago, IL: ODA Books. DOI: 10.13140/RG.2.1.1368.3286] ODA requires no assumptions regarding parent distributions (so _p_ is always exact), and ODA always splines (discretizes) an ordered attribute so as to explicitly maximize the (weighted) classification accuracy.^[Yarnold, P.R., & Soltysik, R.C. (2005). _Optimal Data Analysis: A Guidebook with Software for Windows._ Washington, D.C.: APA Books.]

Data for this trial were entered as tabled frequency counts as shown in **Table 1**. The columns compare Treatment 1 (0) to Treatment 2 (1) according to the number of migrane attacks experienced (rows). A summary of the data from this study follows:
```{r}
library(epitools)

#Summary data, expanded for writing to file, then presented as a cross class table
data <- matrix(c(13,9,4,2,1,1,3,0,5,13,6,1,2,3,3,1),
               ncol=2,nrow=8,dimnames=list(c(0,1,2,3,4,5,6,7),c(0,1)))

data.raw <- setNames(expand.table(data),c("Attacks","Treatment 2"))

data.tab <- table(cbind(data.raw[1],data.raw[2]))

## Table 1: Data for 67 Patients from a Clinical Trial of Two Migraine Treatments
print(data.tab)

```

The MegaODA software was employed via `ODA` to associate the ordered attribute of the number of migrane attacks with treatment arm in the trial. Consistent with Appleton’s spline, the resulting ODA model was:

- _If_ Migrane Attacks=0, _then_ predict Treatment 1; **Otherwise:**
- _If_ Migrane Attacks>0, _then_ predict Treatment 2.

Using the ODA model to classify the sample data yielded the confusion matrix presented in **Table 2**.

```{r}
library(ODA)

#Not run-already exists
#ODAtree("Example 4")

#setwd("~/vignettes/data/Example 4/Runs")

#write.csv(data.raw,file="data3.csv",row.names=F)

#setwd("~/vignettes/data/Example 4")

#ODAclean(data="data4.csv",output=1)

## Load data.csv file and obtain recoded variable names
ODAload(1)
# Recoded variable (v) names to be used in ODA analysis
print(key.1)

## Run MegaODA via ODArun() specifying binary class and ordered attribute variables
#Not run-already exists
#ODArun(run=1,vstart="v1",vend="v2",class="v2",attribute="v1", 
#                                   miss=-9,mcarlo=T,iter=25000)

## Load ODA model into active environment and evaluate output
#ODAparse(1) 
#Message: The default output file name of MODEL.OUT was applied.
#ODA Model summary:
#  Model        IF    THEN       IF    THEN Overall ESS(%) Overall ESP(%) MC P-value 
#      1 V1 <= 0.5  V2 = 0 0.5 < V1  V2 = 1         24.69%         31.41%   0.086760 
#  LOO ESS(%) LOO ESP(%) LOO P-value
#      24.69%     31.41%     .021822

## Display model performance metrics
#print(oda.perf.1)
#Class_variable  Class_name Attribute Attribute_name  PAC  mpac  sens  spec ess.m 
#            V2 Treatment.2        V1        Attacks 62.7 62.34 85.29 39.39 24.68 
# mpv   ppv   npv esp.m   OR  LL.OR UL.OR
# 65.7 59.18 72.22  31.4 3.77  1.16 12.25

## Table 2. Confusion Matrix for a Binary Class, Ordered Attribute, ODA Model
#print(oda.list.1)
#   x
#v2   0  1
#  0 13 20
#  1  5 29

# Here, v2=Observed Treatment 2  status (1 or 0)
# and x=ODA predicted Treatment 2 status (1 or 0)

#Sens for Treatment 1 = 39.4%
round(13/(13+20),3)*100

#Sens for Treatment 2 = 85.3%
round(29/(29+5),3)*100

#PV for Treatment 1 = 72.2%
round(13/(13+5),3)*100

#PV for Treatment 2 = 59.2%
round(29/(29+20),3)*100

```

The model ESS of 24.7 is marginally lower than the minimum criterion for an effect of moderate strength.^4^ The model correctly classified 13 (39%) of 33 patients undergoing Treatment 1, and 29 (85%) of 34 patients undergoing Treatment 2. The model was correct 72% of the time it predicted a patient experienced Treatment 1, and 59% of the time it predicted a patient experienced Treatment 2. For this result, exact two-tailed (i.e., neither treatment was hypothesized _a priori_ to be superior) _p_<0.087. Stable classification performance by ESS was obtained in “leave-one-out” (one-sample jackknife) validity analysis,^4^ and was statistically significant (exact _p_<0.0219) because the ODA model identified in training analysis is applied _a priori_ in jackknife analysis.^4^

